exp2_lik_by_trial

### 2020-08-26 Experiment_2/exp2_lik_by_trial.Rmd
added lik by trial exp 2

---
title: "Bayesian_models_exp2"
output: html_document
---

@@

@@

library(dplyr)

####### test data

# dat <- subset(dat,human_id == "bZg1T3MNoItL6b7qAAAj")
# dat <- subset(dat,human_id == "J3SqvTcFUwXtSC6wAAB0")
# dat <- subset(dat,human_id == "gRRZ0rUagyVejWTOAAFl")
dat <- subset(dat,human_id == "vGnDPRowetREQVmwAADL")
# dat <- subset(dat,human_id == "wi_zWhftr9MdqTmCAAHs")

#######

dat <- as_tibble(dat) %>% group_by(human_id,game)
dat <- dat %>%
  mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

```{r}
## the following computes probabilities of ai actions assuming
## strategy is always followed. (level refers to ai level)

@@

```{r}
# group data by subjective component id

# if human does not distinguish between players, we can run with these predictions as is
# if human does distinguish between players, we need to set the predictions to uniform at the start of a stage

# if human generalizes over games, we can run update the prior generally
# if human generalizes over rounds but not games, we need to reset the prior at the start of a game
# if human does not generalize, we need to reset the prior at the start of each ?stage?

Bayes_model_LL <- function(par,data,distinct_opponent = TRUE, generalize = c("game","stage","no"), return = "-2loglik"){

  # data is a subset for a single participant
  generalize <- match.arg(generalize)
  alpha <- par[1] # probability that ai opponent plays according to strategy
  if(length(par) > 1) {
    beta <- par[2] # probability that human plays according to best response 
  } else{
    beta <- 1 # probability that human plays according to best response 
  }

  prior <- c(1,1,1) # prior alpha for dirichlet on p(level)
  prior <- prior/sum(prior)

  # use alpha to change the "deterministic" predictions
  make_probs <- function(x) {
    (1-alpha)*(1/3) + alpha*x
  }
  dat <- ungroup(data) %>%
    mutate_at(vars(starts_with("pred_")),make_probs)

  if(distinct_opponent) {
    # start of stage should not be predictable
    dat[dat$round == 1,paste0(paste0("pred_",c("a1_","a2_","a3_")),rep(c("level0","level1","level2"),each=3))] <- 1/3
    # group also by opponent
    dat <- group_by(dat,round_condition)
  }

  if(generalize == "stage") {
    # generalize over rounds but not game, so group by game
    dat <- group_by(dat,game,add=TRUE)
  } else if(generalize == "no") {
    # group by game and stage
    dat <- group_by(dat,game,stage,add=TRUE)
  }

  # compute likelihood of ai action
  dat <- dat %>%
    mutate(
      lik_level0 = case_when(
        ai_action_num == 1 ~ pred_a1_level0,
        ai_action_num == 2 ~ pred_a2_level0,
        ai_action_num == 3 ~ pred_a3_level0
      ),
      lik_level1 = case_when(
        ai_action_num == 1 ~ pred_a1_level1,
        ai_action_num == 2 ~ pred_a2_level1,
        ai_action_num == 3 ~ pred_a3_level1
      ),
      lik_level2 = case_when(
        ai_action_num == 1 ~ pred_a1_level2,
        ai_action_num == 2 ~ pred_a2_level2,
        ai_action_num == 3 ~ pred_a3_level2
      )
    )

  # use likelihood to compute the posterior predictive probability of each level
  dat <- dat %>%
    mutate(logpost_level0 = lag(log(prior[1]) + cumsum(log(lik_level0)),default=log(prior[1])),
           logpost_level1 = lag(log(prior[2]) + cumsum(log(lik_level1)),default=log(prior[2])),
           logpost_level2 = lag(log(prior[3]) + cumsum(log(lik_level2)),default=log(prior[3]))) %>%
    # you can add or subtract any constant from the log-likelihoods; this can aid in precision
    mutate(min = pmin(logpost_level0,logpost_level1, logpost_level2)) %>%
      mutate(normalize = exp(logpost_level0 - min) + exp(logpost_level1 - min) + exp(logpost_level2 - min)) %>%
        mutate(post_level0 = exp(logpost_level0 - min)/normalize,
               post_level1 = exp(logpost_level1 - min)/normalize,
               post_level2 = exp(logpost_level2 - min)/normalize)

    # cat(as.character(c(dat$post_level0, dat$post_level1, dat$post_level2)), "\n")

  # use posterior predictive probability to predict probability of each ai and then human action
  # opp is ai and self is human player here
  dat <- dat %>%
    mutate(p_opp_a1 = post_level0*pred_a1_level0 + post_level1*pred_a1_level1 + post_level2*pred_a1_level2,
           p_opp_a2 = post_level0*pred_a2_level0 + post_level1*pred_a2_level1 + post_level2*pred_a2_level2,
           p_opp_a3 = post_level0*pred_a3_level0 + post_level1*pred_a3_level1 + post_level2*pred_a3_level2
    ) %>%
      mutate(p_self_a1 = case_when(
              game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a2 + .5*p_opp_a3),
              TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a3)
                ),
             p_self_a2 = case_when(
               game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a1 + .5*p_opp_a3),
               TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a1)
             ),
             p_self_a3 = case_when(
               game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a1 + .5*p_opp_a2),
               TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a2)
             )
        )

  # now finally compute the likelihood of human actions 
  dat <- dat %>% 
    mutate(
      loglik = case_when(
        human_action_num == 1 ~ log(p_self_a1),
        human_action_num == 2 ~ log(p_self_a2),
        human_action_num == 3 ~ log(p_self_a3))
    )

  #cat(as.character(exp(dat$loglik)), "\n", "\n")
  if (return == "-2loglik") {
    ret <- -2*sum(dat$loglik)
    if(is.infinite(ret) || is.nan(ret)) {
      return(1e+300)
    } else {
    return(ret) 
    } 
  } else if (return == "likelihood_by_trial") {
    return(exp(dat$loglik))
  }
}

@@

```{r}
Bayes_model_LL(0.5,data = dat,distinct_opponent = TRUE, generalize = "game", return = "likelihood_by_trial")

#PLot likelihood for first participant as function of parameter 
# likelihoods <- list()
# for (t in 1:99) {
#   likelihoods[t] <- Bayes_model_LL(t/100,data = dat,distinct_opponent = TRUE, generalize = "game")
# }
# 
# jpeg(file="vGnDPRowetREQVmwAADL.jpeg")
# plot(unlist(likelihoods))
# dev.off()

```{r, ECHO = FALSE}

library(DEoptim)

Bayes_same_game <- Bayes_same_stage <- Bayes_same_no <- 
  Bayes_distinct_game <- Bayes_distinct_stage <- Bayes_distinct_no <- 
    list()

# USing DEOptim

for(id in levels(dat$human_id)) {
  tdat <- subset(dat,human_id == id)
  ctrl <- DEoptim.control(NP = 20, itermax=50)
  Bayes_same_game[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                   distinct_opponent = FALSE, generalize = "game",control=ctrl)
  Bayes_same_stage[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                   distinct_opponent = FALSE, generalize = "stage",control=ctrl)
  Bayes_same_no[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                   distinct_opponent = FALSE, generalize = "no",control=ctrl)
  Bayes_distinct_game[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                   distinct_opponent = TRUE, generalize = "game",control=ctrl)
  Bayes_distinct_stage[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                        distinct_opponent = TRUE, generalize = "stage",control=ctrl)
  Bayes_distinct_no[[id]] <- DEoptim(Bayes_model_LL, lower=c(0), upper = c(1), data = tdat,
                                 distinct_opponent = TRUE, generalize = "no",control=ctrl)
}

# using optim

# for(id in levels(dat$human_id)) {
#   tdat <- subset(dat,human_id == id)
#   # ctrl <- DEoptim.control(NP = 20, itermax=50)
#   
#   Bayes_same_game[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "game", lower = c(0), upper = c(1),method="L-BFGS-B")
#   
#   
#   Bayes_same_stage[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "stage", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_same_no[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = FALSE, 
#                                  generalize = "no", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_game[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "game", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_stage[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "stage", lower = c(0), upper = c(1),method="L-BFGS-B")
# 
#   Bayes_distinct_no[[id]] <- optim(c(0.1),fn=Bayes_model_LL,data=tdat, distinct_opponent = TRUE, 
#                                  generalize = "no", lower = c(0), upper = c(1),method="L-BFGS-B")
# }
```

```{r}
# this will implement a pseudo-multinomial-dirichlet
pBayes_model_LL <- function(par,data,distinct_opponent = TRUE, generalize = c("game","stage","no")) {
  generalize <- match.arg(generalize)
  alpha <- par[1] # probability that opponent plays according to strategy
  if(length(par) > 1) {
    gamma <- par[2] # probability that human plays according to best response 
  } else{
    gamma <- 1 # probability that human plays according to best response 
  }
  if(length(par) > 2) {
    beta <- par[3] # probability that human plays according to best response 
  } else{
    beta <- 1 # probability that human plays according to best response 
  }

  prior <- c(gamma,gamma,gamma) # prior alpha for dirichlet on p(level)
  #prior <- prior/sum(prior)

  # use alpha to change the "deterministic" predictions
  make_probs <- function(x) {
    (1-alpha)*(1/3) + alpha*x
  }
  dat <- ungroup(data) %>%
    mutate_at(vars(starts_with("pred_")),make_probs)

  if(distinct_opponent) {
    # start of round should not be predictable
    dat[dat$round == 1,paste0(paste0("pred_",c("a1_","a2_","a3_")),rep(c("level0","level1","level2"),each=3))] <- 1/3
    # group also by opponent
    dat <- group_by(dat,opp_type)
  }

  if(generalize == "stage") {
    # generalize over rounds but not game, so group by game
    dat <- group_by(dat,game,add=TRUE)
  } else if(generalize == "no") {
    # group by game and stage
    dat <- group_by(dat,game,stage,add=TRUE)
  }

  # compute likelihood of ai action
  dat <- dat %>%
    mutate(
      lik_level0 = case_when(
        ai_action_num == 1 ~ pred_a1_level0,
        ai_action_num == 2 ~ pred_a2_level0,
        ai_action_num == 3 ~ pred_a3_level0
      ),
      lik_level1 = case_when(
        ai_action_num == 1 ~ pred_a1_level1,
        ai_action_num == 2 ~ pred_a2_level1,
        ai_action_num == 3 ~ pred_a3_level1
      ),
      lik_level2 = case_when(
        ai_action_num == 1 ~ pred_a1_level2,
        ai_action_num == 2 ~ pred_a2_level2,
        ai_action_num == 3 ~ pred_a3_level2
      )
    )

  # normalize the likeihoods to give a pseudo-observation of strategy
  dat <- dat %>%
    # compute normalizing factor
    mutate(normalize = lik_level0 + lik_level1 + lik_level2) %>%
    # avoid division by 0
    mutate(normalize = replace(normalize, normalize == 0, 1)) %>%
    # normalize
    mutate(pseudo_obs_level0 = lik_level0/normalize,
           pseudo_obs_level1 = lik_level1/normalize,
           pseudo_obs_level2 = lik_level2/normalize)

  # use pseudo observations to compute the posterior predictive probability of each level
  dat <- dat %>%
    mutate(post_level0 = lag(prior[1] + cumsum(pseudo_obs_level0),default=prior[1]),
           post_level1 = lag(prior[2] + cumsum(pseudo_obs_level1),default=prior[2]),
           post_level2 = lag(prior[3] + cumsum(pseudo_obs_level2),default=prior[3])) %>%
      mutate(normalize = post_level0 + post_level1 + post_level2) %>%
        mutate(post_level0 = post_level0/normalize,
               post_level1 = post_level1/normalize,
               post_level2 = post_level2/normalize)

  # use posterior predictive probability to predict probability of each ai and then human action
  dat <- dat %>%
    mutate(p_opp_a1 = post_level0*pred_a1_level0 + post_level1*pred_a1_level1 + post_level2*pred_a1_level2,
           p_opp_a2 = post_level0*pred_a2_level0 + post_level1*pred_a2_level1 + post_level2*pred_a2_level2,
           p_opp_a3 = post_level0*pred_a3_level0 + post_level1*pred_a3_level1 + post_level2*pred_a3_level2
    ) %>%
    mutate(p_self_a1 = case_when(
      game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a2 + .5*p_opp_a3),
      TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a3)
    ),
    p_self_a2 = case_when(
      game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a1 + .5*p_opp_a3),
      TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a1)
    ),
    p_self_a3 = case_when(
      game == "shootout" ~ (1-beta)*(1/3) + beta*(.5*p_opp_a1 + .5*p_opp_a2),
      TRUE ~ (1-beta)*(1/3) + beta*(p_opp_a2)
    )
    )

  # now finally compute the likelihood of human actions 
  dat <- dat %>% 
    mutate(
      loglik = case_when(
        human_action_num == 1 ~ log(p_self_a1),
        human_action_num == 2 ~ log(p_self_a2),
        human_action_num == 3 ~ log(p_self_a3))
    )

  return(-2*sum(dat$loglik))
}

```{r}
pBayes_same_game <- pBayes_same_stage <- pBayes_same_no <- 
  pBayes_distinct_game <- pBayes_distinct_stage <- pBayes_distinct_no <- 
  list()

for(id in levels(dat$human_id)) {
  tdat <- subset(dat,human_id == id)
  ctrl <- DEoptim.control(NP = 20, itermax=50)
  pBayes_same_game[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                   distinct_opponent = FALSE, generalize = "game",control=ctrl)
  pBayes_same_stage[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                    distinct_opponent = FALSE, generalize = "stage",control=ctrl)
  pBayes_same_no[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                 distinct_opponent = FALSE, generalize = "no",control=ctrl)
  pBayes_distinct_game[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                       distinct_opponent = TRUE, generalize = "game",control=ctrl)
  pBayes_distinct_stage[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                        distinct_opponent = TRUE, generalize = "stage",control=ctrl)
  pBayes_distinct_no[[id]] <- DEoptim(pBayes_model_LL, lower=c(0.1,1), upper = c(1,100), data = tdat,
                                     distinct_opponent = TRUE, generalize = "no",control=ctrl)
}
```

>>>>

---
title: "exp2_lik_by_trial"
output: html_document
---

@@

```{r}
library(dplyr)
library(tidyr)

exp2_optim_results <- read.csv("exp2_results.csv")

@@

# Preparing data for running functions to calculate likelihoods

# dat <- as_tibble(dat) %>% group_by(human_id,game)
  dat <- dat %>%
   mutate(ai_action_prev = lag(ai_action,1), human_action_prev = lag(human_action,1))

@@

lik_by_trial_2 <- data.frame()

for(id in unique(dat$human_id)) {

  tdat <- dat %>% subset(human_id == id)

  # QLS likelihoods 
  pars_QLS <- unlist(filter(exp2_optim_results, ID == id)%>%select(QL_states_alpha,QL_states_Beta))
  tdat$QLS_lik <-  exp2_QL_states(pars_QLS,tdat,"likelihood_by_trial",0.0, FALSE)

  # QLS within Transfer likelihoods 
  pars_QLS_within <- unlist(filter(exp2_optim_results, ID == id)%>%select(QLS_within_Tr_alpha,QLS_within_Tr_Beta))
  tdat$QLS_lik_within <-  exp2_QL_states(pars_QLS_within,tdat,"likelihood_by_trial",gamma = 0, TRUE)
  #                
  # 
  # Bayes model no transfer likelihoods
  theta_NT <- unlist(filter(exp2_optim_results, ID == id)%>%select(theta_no_tr))
  tdat$Bayes_NT_lik <- Bayes_model_LL(theta_NT, tdat, distinct_opponent = TRUE, generalize = "no",return = "likelihood_by_trial")

  # Bayes model with Transfer likelihood
  theta_Tr <- unlist(filter(exp2_optim_results, ID == id)%>%select(theta_transfer))
  tdat$Bayes_Tr_lik <- Bayes_model_LL(theta_Tr, tdat, distinct_opponent = TRUE, generalize = "game",return = "likelihood_by_trial")

  lik_by_trial_2 <- rbind(lik_by_trial_2,tdat)
}
```

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

mean_lik_2 <- lik_by_trial_2 %>%
  select(human_id,game,stage,round_condition, round, QLS_lik, QLS_lik_within, Bayes_NT_lik, Bayes_Tr_lik) %>%
  group_by(game,stage,round) %>%
  summarise(mean_QLS_lik = mean(QLS_lik, na.rm = TRUE),
            mean_QLS_within_lik = mean(QLS_lik_within, na.rm = TRUE),
            mean_NT_lik = mean(Bayes_NT_lik, na.rm = TRUE),
            mean_Tr_lik = mean(Bayes_Tr_lik, na.rm = TRUE))

mean_lik$game <- factor(mean_lik$game,levels=c("rps","fwg","shootout"))

data_long <- gather(mean_lik_2, strategy, probability, mean_QLS_lik:mean_Tr_lik, factor_key=TRUE)

# make sure the different games are ordered in the way they were played
data_long$game <- factor(data_long$game,levels=c("rps","fwg","shootout"))
ggplot(data_long,aes(x=round,y=probability,colour=strategy)) + geom_line() + facet_grid(game~stage)

@@
